### 2018-03-16 paper/4-build.tex
4-build

\subsubsection{Topological}

@@

\subsubsection{Reordering}

@@

This requires a way to abort tasks that have failed due to out-of-date
dependencies. It is also not minimal in the sense that a task may start, do some
meaningful work, then abort. However, in the case of an \hs{Applicative} system,
that work is zero.

\subsubsection{Recursive}

@@

This requires a way to suspend a running task, which can be done with cheap
green threads and blocking (the original approach of \Shake) or with a
continuation-passing style (what \Shake does now). An alternative approach to
suspending a task is to abort it and restart it again later, at the cost
of doing additional work.

@@

With a dirty bit it is possible to achieve minimality. To achieve early cutoff
(\S\ref{sec-background-shake}) it would be important to not set the dirty bit
after a computation that did not change the value. \Excel could use this
approach, but does not. In contrast, \Make cannot implement early cutoff -- to
do so it would have to mark the node clean (so it would not rebuild in the next
run) and at the same time not mark the things it depends on dirty -- an
impossible task with only the ability to update to the latest modification time.

\todo{NM}{I stopped here, as the traces seem not finalised yet.}

@@

We assume that \hs{Hash v} is a small constant size, constructed from hashing the
underlying \hs{v} rather than storing it directly. Checking a trace requires
ensuring all the dependencies are up to date (using whatever ordering strategy
as per \S\ref{sec-dependency-orderings}), then comparing if the dependencies are
same as the current value and that the result is the same. After computing a
fresh value we add its \hs{Trace k v} to the list. Therefore the information stored
by a build system that verifies traces is \hs{[Trace k v]}.

\todo{NM}{We need to emphasise that lists are convenient for expressing the semantics,
but actual implementation uses a data structure optimised for a specific build system.}

@@

\todo{NM}{Clarify why the indirection (first get the hash, then look it up)?}

\begin{minted}{haskell}
data Traces k v = Traces
    { traces :: [Trace k v]
    , cache  :: Map Hash v }
\end{minted}

We have a list of traces, plus a \hs{Map} from the hash to the actual contents. Checking a trace is the same as before, but if the result is the only thing that is different we can simply retrieve a fresh result from \hs{cache} \emph{without} recomputing it.

For both verifying and constructive choices it is possible to choose smarter data structures than simply a list of \hs{Trace}, which we discuss in \S\ref{sec-smart-traces}.

\subsection{Summary}\label{sec-design-space}

\todo{NM}{Check the table and add a decription.}

\begin{table}[h]
\smaller
\centering
\begin{tabular}{l||c|c|c}
\hline
Property           & Topological & Reordering & Recursive    \\\hline
\hline
Dirty bit          & \Make                   & \Excel                   & \Shake without early cutoff \\\hline
Verifying trace    & \Make with early cutoff & \Excel with early cutoff & \Shake                      \\\hline
Constructive trace & \Bazel                  & Cloud \Excel             & Cloud \Shake                \\\hline
\end{tabular}
\vspace{0.5mm}
\caption{Build systems \`a la carte.\label{tab-build-systems}}
\end{table}

>>>>

\subsubsection{Topological}\label{sec-topological}

@@

\subsubsection{Reordering}\label{sec-reordering}

@@

This approach requires a way to abort tasks that have failed due to out-of-date
dependencies. It is also not minimal in the sense that a task may start, do some
meaningful work, then abort. However, in the case of an \hs{Applicative} system,
that work is zero.

\subsubsection{Recursive}\label{sec-recursive}

@@

This approach requires a way to suspend a running task, which can be done with cheap
green threads and blocking (the original approach of \Shake) or with a
continuation-passing style (what \Shake does now). An alternative approach to
suspending a task is to abort it and restart it again later, at the cost
of doing additional work.

@@

When using a dirty bit, it is necessary to check all the dependencies of a key. For applicative build systems that list is simple to obtain, but for monadic build systems there is no general way to get all dependencies. Instead \Excel computes a \textit{static approximation} of the dependencies. For applicative rules that approximation is correct. For functions such as \textsf{IF} it marks the cell dirty if \textit{any} potential dependency has changed, even on the untaken \textif{if} branch. For functions such as \textsf{INDIRECT} whose dependencies cannot be guessed, it conservatively assumes the dependencies have always changed.

With a dirty bit it is simple to achieve minimality. However, to achieve early cutoff
(\S\ref{sec-background-shake}) it would be important to not set the dirty bit
after a computation that did not change the value. \Excel could use this
approach, but does not. In contrast, \Make cannot implement early cutoff -- to
do so it would have to mark the node clean (so it would not rebuild in the next
run) and at the same time not mark the things it depends on dirty -- an
impossible task with only the ability to update to the latest modification time.

@@

We assume that \hs{Hash v} is a small constant size, constructed from hashing the
underlying \hs{v} rather than storing it directly. Checking a trace requires
ensuring all the dependencies are up to date (using whatever ordering strategy
as per \S\ref{sec-dependency-orderings}), then comparing if the dependencies are
same as the current value and that the result is the same.

A verifying trace build system needs to maintain a set of traces. After computing a
fresh value we add its \hs{Trace k v} to the set. Therefore the information stored
by a build system that verifies traces can be stored as a list or set - we chose \hs{[Trace k v]} for simplicity. In practice, different build systems can optimise the data structures used by traces for their specific use cases, which we discuss in \S\ref{sec-smart-traces}.

@@

\begin{minted}{haskell}
data Traces k v = Traces
    { traces :: [Trace k v]
    , cache  :: Map (Hash v) v }
\end{minted}

We have a list of traces, plus a \hs{Map} from the hash to the actual contents. Checking a trace is the same as before, but if the result is the only thing that is different we can simply retrieve a fresh result from \hs{cache} \emph{without} recomputing it. We split the traces and caches as in cloud interactions to a remote server the checking system may have to examine many traces/hashes, but only at most one complete file.

\subsection{Build Systems \`a la Carte}\label{sec-design-space}

\begin{table}[h]
\smaller
\centering
\begin{tabular}{l||c|c|c}
\hline
Property           & Topological \S\ref{sec-topological} & Reordering \S\ref{sec-reordering} & Recursive \S\ref{sec-recursive}    \\\hline
\hline
Dirty bit \S\ref{sec-dirty-bit}                    & \Make    & \Excel                   & Approximate \Shake  \\\hline
Verifying trace \S\ref{sec-verifying-traces}       & \Ninja   & Traced \Excel            & \Shake                      \\\hline
Constructive trace \S\ref{sec-constructive-traces} & \Bazel   & Cloud \Excel             & Cloud \Shake                \\\hline
\end{tabular}
\vspace{0.5mm}
\caption{Build systems \`a la Carte.\label{tab-build-systems}}
\end{table}

With the information in this section we can build a table comparing the dependency order strategy with the out-of-date keys strategy, providing 9 possible build systems, 5 of which are actually inhabited by existing build systems (we discuss \Ninja \cite{ninja} in \S\ref{sec-related-build}). Of the remaining 4 spots, we believe neither Traced or Cloud \Excel make sense - the \Excel approach of reordering combined with static approximations reduces the memory usage significantly. However, as soon as you are paying the cost of storing traces, that benefit is gone. The advantage of Approximate \Shake over \Shake would be that it could avoid storing traces and having a separate information database, but that advantage is minor compared to the technical restrictions and approximations it would provide, so we consider it unlikely to be built. The Cloud \Shake system seems like an interesting and important point in the design space, which we explore further in \S\ref{sec-implementation-cloud-shake}.
